Solution of Differential Algebraic Equations with Manifold Constraints on the Latent Space of an Autoencoder

### 2019-05-14 paper/paper.tex
writing more

Representing complex physical systems with multiple speicies, phases,
chemical reactions, and heterogenous scale effects poses an
intractability problem to modeling. How do we analyze complex
experimental data and \emph{ab initio} simulations and derive new
macroscale relations to describe the systems? How do we then implement
these relations in the rest of our analyses and production-scale
computer simulations? A single material with multiple phase transitions
is a challenge in of itself, and the descriptions become combinatorially
more complex as species and phases are added. This results in
combinatorially more human labor and combinatorially more code which
tends to involve complex branching logic. The descriptions then pose
numerical challenges when incorporated into macroscale simulations due
to obstacles such as discrete phase state machines and primary variable
switching.

Motivations for this work are subsurface energy systems such as shale
and oceanic natural gas reservoirs in shales or engineered geothermal
systems.

This work is only considered with the "single beaker" problem; the flow
and transport aspects have similar challenges. (The next stage of this
work will "connect the beakers".)

This work proposes a new methodology that steps away from the
traditional approaches

In this approach, we directly \emph{look for solvable representations}
instead of representations that carry human intuition. We value
simulation runtime and robustness. Further, we must remember the value
of the human effort involved: beyond improving computational aspects,
we are fully automating part of the human involvement in programming
the simulator.

@@

Consider a general differential algebraic equation of the form

\begin{equation}
z(\dot{u},u) = 0
\end{equation}

We are interested in equations that have the form

\begin{align}
\frac{\mathrm{d}u}{\mathrm{d}t} & = v \\
0 & = c(u,v)
\end{align}

Where do such equations arise? A simple example is the pendulum, which
will be explored in Section 2.

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \\
m \dot{x} & = f x/L \\
m \dot{y} & = f y/L + m g \\
x^2 + y^2 & = L^2
\end{align}

@@

We also consider the phase equilibria; at a larger scale it is possible
for a control volume to be a mixture of two phase coexisting, such as
boiling with some mixture saturation \(S_{gas}=1-S_{liq}\).

@@

The complicatation is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for water.
The material can exist on this section of the surface as a mixture of
the two phases.

@@

We will also note right now that applicaitons also involve continuum
mechanics and chemical reactions! The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve.

@@

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/phase_diagram.png}
\caption{}
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/water_eos.png}
\caption{}
\end{figure}

@@

An archetypical constraint problem the pendulum. Posed as dynamics, it
has three unknowns wtih two second-order equations and a constraint,

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \\
m \ddot{x} & = f x/L \\
m \ddot{y} & = f y/L + m g \\
x^2 + y^2 & = L^2
\end{align}

In our methodology, we do not have an explicit equation for the
constraint, but a dataset of (x,y) pairs that are on the manifold. For
our proof-of-concept test, we manufacture this data for the pendulum.
Note that we are considering the nonlinear pendulum.

There are many ways to solve this cannonical equation: Lagrange
multipliers (\(f(t)\) is the multiplier in the way we have written it
now), penalties, change of variables, etc. The change of variables
method is to introduce a new variable \(\theta\) that parameterizes
\(x=L\cos\theta\) and \(y=L\sin\theta\):

\begin{align}
\text{Solve for}\, \theta(t) \text{and} \, f(t) \, \text{satisfying:} \\
m \ddot{x}(\theta) & = f x(\theta)/L \\
m \ddot{y}(\theta) & = f y(\theta)/L + m g \\
\end{align}

In this example, this lets us collapse the problem into a
single-component Ordinary Differential Equation.

The manifold for this simple problem is smooth, so two nested
polynomials forms the autoencoder:

\[\left\{\begin{array}{c} x\\y\end{array}\right\}
\rightarrow W^1 x^n \rightarrow q \rightarrow W^2 x^n \rightarrow 
\left\{\begin{array}{c} x\\y\end{array}\right\}\]

(We explicitly write the bias \(+b\); it is equivalent to including
\(x^0=1\) in the polynomial set.) The Lagrangian for the pendulum is

\begin{equation}
L = \frac{1}{2}m\left(\dot{x}^2 + \dot{y}^2\right) - m g y
\end{equation}

with the constraint that \(x\) and \(y\) lie on the path. The equations
of motion given some parameter space that satisfies the constraint is

>>>>

Representing complex physical systems with multiple speicies, phases,
chemical reactions, and heterogenous scale effects poses an
intractability problem to modeling. How do we analyze complex
experimental data and \emph{ab initio} simulations and derive new
macroscale relations to describe the systems? How do we then implement
these relations in the rest of our analyses and production-scale
computer simulations?

Accurately simulating multispecies reaction and transport with
undergoing phase transitions is a critical aspect to a variety of
problems.
Motivations for this work are subsurface energy systems such as shale
and oceanic natural gas reservoirs in shales or engineered geothermal
systems.

A single material with multiple phase transitions
is a challenge in of itself, and the descriptions become combinatorially
more complex as species and phases are added. This results in
combinatorially more human labor and combinatorially more code which
tends to involve complex branching logic. The descriptions then pose
numerical challenges when incorporated into macroscale simulations due
to obstacles such as discrete phase state machines and primary variable
switching.

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot flux_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot flux_{CO_2} + s \\
  \partial_t e& = \nabla \cdot flux_{heat} + s
\end{align}
More species add more rows, and applicaitons also involve continuum
mechanics and chemical reactions! The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{my millstone paper}

This work only considers the "single beaker" problem; the flow
and transport aspects have similar challenges. (The next stage of this
work will "connect the beakers".)

In section 2, we will cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \leftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
We will use this set of possible values viewpoint to train an
autoencoder on the set of $u$ for a latent space parameterized by $q$:
\begin{equation}
u = D(q), \; q=E(u) \;\forall u \in U
\end{equation}
which satisifies
\begin{equation}
  D(q)\in U \forall q.
\end{equation}
We can then write the original equation by
\begin{equation}
z(q,\partial_t q)=0
\end{equation}

@@

This work proposes a new methodology that steps away from the
traditional approaches. 
In this approach, we directly \emph{look for solvable representations}
instead of representations that carry human intuition. We value
simulation runtime and robustness. Further, we must remember the value
of the human effort involved: beyond improving computational aspects,
we are fully automating part of the human involvement in programming
the simulator.

In section 2, the balance of mass and energy will be described for a
single multiphase material. We will discuss the current state of the
art algorithm for deriving relationships for multiphase materials, and
then the primary-variable switching algorithm that is used in
state-of-the art solvers. The equation will then be recast as a
constraint problem in th above form.
In section 3, we will demonstrate applying the autoencoder learning
methodology to solve the pendulum,
which can be considered the archetypical constraint problem.
In section 4, the autoencoder methodology wil lbe applied to the
equation of state of water. In section 5, the new simulation process
will be described, wherein the learned decoders are placed into the
balance laws, forming newly discovered equations on learned
representations. The software implementation of the training and
simulation programs are detailed in section 6. The new simulation
framework on sets of learned representations is evaluated on
multiple datasets of varying extents of water with a
hyperparameter search on a suite of test problems to prouce the discussion of section 7, as
well as a systematic automated testing procedure of a
machine-dominated programming methodology.
Active research and software development directions are
concluded in the final section.

The notebooks and software implementations are released open source to
accompany this work. They can be found at the github repository
located at: \href{https://github.com/afqueiruga/latentsim}
Certain components are factored into submodule repositories which are
each released under the LGPL at seperate repositories referenced in
the implementation details.

@@

@@

The complicatation is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.

We also consider the phase equilibria; at a larger scale it is possible
for a control volume to be a mixture of two phase coexisting, such as
boiling with some mixture saturation \(S_{gas}=1-S_{liq}\).
(For a very fine scale simulation, bubbles would be explicitly modeled.)
The material can exist on this section of the surface as a mixture of
the two phases.

@@

@@

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/phase_diagram.png}
\caption{Phase boundaries for water }
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/figures/water_eos.png}
\caption{\label{fig:pTrho} The $\log{p}$, $T$, $\rho$ equation of
  state surface for water. This plot spans four phases, three
  equilibria, and the triple point, which are each given their own color.
  There is a fourth dimension to the surface of enthalpy $h$ that exhibits
  similar discontinuities; the $\log{p}$, $T$, $h$ surface is plotted
  in the Appendix.}
\end{figure}

There are even more known phases of crystal ice which for water
For water, the supercritical phase and exotic ices only occur in
laboratory conditions or naturally near the
crust \cite{}. For other materials these exotic phases can be
applicable. For example, in geoengineered systems using $CO_2$
such as geothermal plants are carbon sequestration facilities, the
working fluid is usually in the supercritical regime at depth.

@@

An archetypical constraint problem the pendulum. Posed as dynamics
problem with a Lagrange-multipler-esque centripetal force, it
has three unknowns wtih two second-order equations and a constraint,

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
x^2 + y^2 & = R^2
\end{align}

Note that we are considering the nonlinear pendulum.

There are many ways to solve this cannonical equation: Lagrange
multipliers (\(f(t)\) is the multiplier in the way we have written it
above), penalties, change of variables, etc. The change of variables
method exploits the Lagrangian mechanics formulation, where we state
the Lagrangian,
\begin{equation}
L = \frac{1}{2}m\left(\dot{x}^2 + \dot{y}^2\right) - m g y
\end{equation}
with the constraint that \(x\) and \(y\) lie on the path.
We can introduce a new variable \(\theta\) that parameterizes
\(x=R\cos\theta\) and \(y=R\sin\theta\) and rederive the equations of
motion with the familiar,
\begin{align}
\text{Solve for}\, \theta(t) \ \text{satisfying:} \\
\frac{\mathrm{d}}{\mathrm{d}t} \frac{\partial L}{\partial
  \dot{\theta}} =
  \frac{\partial L}{\partial \theta}
\end{align}
In this example, this lets us collapse the problem into a
single-component second-roder differential equation.

In our methodology, we do not have an explicit equation for the
constraint, but a dataset of $X$ of $(x,y)$ pairs that are on the manifold. For
our proof-of-concept test, we manufacture this data for the circular
pendulum to verify against the analytical solution:
\begin{equation}
X = \left\{ (R\cos\theta_i, R\sin\theta_i)\; | \;\theta_i \in [-5\pi/4,\pi/4] \right\}
\end{equation}
We also try a ``goofier'' looking curve with no known closed form
solution to exercise the methodology,

(For the equation of state problem, this is analogous to the dataset
of $p,T,\rho,h$ 4-tuples that make the plot in Figure \ref{fig:pTrho}.)

The manifold for this simple problem is smooth, so two nested
polynomials forms the autoencoder:
\[\left\{\begin{array}{c} x\\y\end{array}\right\}
\rightarrow W_{enc} x^n \rightarrow q \rightarrow W_{dec} x^n \rightarrow 
\left\{\begin{array}{c} x\\y\end{array}\right\}\]

(We explicitly write the bias \(+b\); it is equivalent to including
\(x^0=1\) in the polynomial set.)
The equations
of motion given some parameter space that satisfies the constraint is
